RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2016-06-27 manuscripts/manuscript.tex
omit rrdf

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{RNeXML: a package for reading and writing richly annotated phylogenetic,
character, and trait data in R}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[ropensci]{Scott Chamberlain}

@@

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[ropensci]{University of California, Berkeley, CA, USA}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}    
    \address[dukeplus]{Center for Genomic and Computational Biology, Duke University, and
National Evolutionary Synthesis Center, Durham, NC, USA}

@@

The exchange problem for phylogenetic data is particularly acute in
light of the challenges in finding and sharing phylogenetic data without
the otherwise common loss of most data and metadata semantics (Stoltzfus
\emph{et al.} 2012; Drew \emph{et al.} 2013; Cranston \emph{et al.}
2014). For example, the still popular NEXUS file format (Maddison
\emph{et al.} 1997) cannot consistently represent horizontal gene
transfer or ambiguity in reading a character (such as a DNA sequence
base pair). This and other limitations have led to modifications of
NEXUS in different ways for different needs, with the unfortunate result
that NEXUS files generated by one program can be incompatible with
another (Vos \emph{et al.} 2012). Without a formal grammar, software
based on NEXUS files may also make inconsistent assumptions about
tokens, quoting, or element lengths. Vos et al. (2012) estimates that as
many as 15\% of the NEXUS files in the CIPRES portal contain
unrecoverable but hard to diagnose errors.

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in Vos et al. (2012). In brief, NeXML
was designed with the following important properties. First, NeXML is
defined by a precise grammar that can be programmatically
\textbf{validated}; i.e., it can be verified whether a file precisely
follows this grammar, and therefore whether it can be read (parsed)
without errors by software that uses the NeXML grammar (e.g.~RNeXML) is
predictable. Second, NeXML is \textbf{extensible}: a user can define
representations of new, previously unanticipated information (as we will
illustrate) without violating its defining grammar. Third and most
importantly, NeXML is rich in \textbf{computable semantics}: it is
designed for expressing metadata such that machines can understand their
meaning and make inferences from it. For example, OTUs in a tree or
character matrix for frog species can be linked to concepts in a
formally defined hierarchy of taxonomic concepts such as the Vertebrate
Taxonomy Ontology (Midford \emph{et al.} 2013), which enables a machine
to infer that a query for amphibia is to include the frog data in what
is returned. (For a more broader discussion of the value of such
capabilities for evolutionary and biodiversity science we refer the
reader to Parr et al. (2011).)

@@

The \texttt{RNeXML} package is written entirely in R and available under
a Creative Commons Zero public domain waiver. The current development
version can be found on Github at
\href{}{\url{https://github.com/ropensci/RNeXML}}, and the stable
version can be installed from the CRAN repository. \texttt{RNeXML} is
part of the rOpenSci project. Users of \texttt{RNeXML} are encouraged to
submit bug reports or feature requests in the issues log on Github, or
the phylogenetics R users group list at
\texttt{r-sig-phylo@r-project.org} for help. Vignettes with more
detailed examples of specific features of RNeXML are distributed with
the R package and serve as a supplement to this manuscript. Each of the
vignettes can be found at
\href{}{\url{http://ropensci.github.io/RNeXML/}}.

@@

\begin{verbatim}
         log snout-vent length reef-dwelling
taxon_8             -3.2777799             0
taxon_9              2.0959433             1
taxon_10             3.1373971             0
taxon_1              4.7532824             1
taxon_2             -2.7624146             0
taxon_3              2.1049413             0
taxon_4             -4.9504770             0
taxon_5              1.2714718             1
taxon_6              6.2593966             1
taxon_7              0.9099634             1
\end{verbatim}

@@

\subsection{Creating and populating \texttt{nexml}
objects}\label{creating-and-populating-nexml-objects}

@@

As described in Vos et al. (2012), to encode metadata annotations NeXML
uses the ``Resource Description Framework in Annotations'' (RDFa)
(Prud'hommeaux 2014). This standard provides for a strict
machine-readable format yet enables future backwards compatibility with
compliant NeXML parsers (and thus \texttt{RNeXML}), because the capacity
of a tool to \emph{parse} annotations is not predicated on
\emph{understanding} the meaning of annotations it has not seen before.

@@

Third, \texttt{RNeXML} integrates with the R package \texttt{taxize}
(Chamberlain \& Sz{ö}cs 2013) to mitigate one of the most common
obstacles to reuse of phylogenetic data, namely the misspellings and
inconsistent taxonomic naming with which OTU labels are often fraught.
The \texttt{taxize\_nexml()} method in \texttt{RNeXML} uses
\texttt{taxize} to match OTU labels against the NCBI database, and,
where a unique match is found, it annotates the respective OTU with the
matching NCBI identifier.

@@

Because NeXML expresses all metadata using the RDF standard, and stores
them compliant with RDFa, they can be extracted as an RDF graph,
queried, analyzed, and mashed up with other RDF data, local or on the
web, using a wealth of off-the-shelf tools for working with RDF (see
Prud'hommeaux (2014) or Hartig (2012)). Examples for these possibilities
are included in the \texttt{RNeXML} SPARQL vignette (a recursive acronym
for SPARQL Protocol and RDF Query Language, see
\url{http://www.w3.org/TR/rdf-sparql-query/}), and the package also
comes with a demonstration that can be run from R using the following
command: \texttt{demo("sparql", "RNeXML")}).

@@

Data archiving is increasingly required by scientific journals,
including in evolutionary biology, ecology, and biodiversity (e.g.
Rausher et al. (2010)). The effort involved with preparing and
submitting properly annotated data to archives remains a notable barrier
to the broad adoption of data archiving and sharing as a normal part of
the scholarly publication workflow (Tenopir \emph{et al.} 2011; Stodden
2014). In particular, the majority of phylogenetic trees published in
the scholarly record are inaccessible or lost to the research community
(Drew \emph{et al.} 2013).

@@

Boettiger, C., Vos, R., Chamberlain, S. \& Lapp, H. (2014). RNeXML
v2.0.0. Retrieved from \url{http://dx.doi.org/10.5281/zenodo.13131}

Bollback, J. (2006).\emph{BMC Bioinformatics}, \textbf{7}, 88. Retrieved
from \url{http://dx.doi.org/10.1186/1471-2105-7-88}

Chamberlain, S.A. \& Sz{ö}cs, E. (2013). Taxize: Taxonomic search and
retrieval in r. \emph{F1000Research}. Retrieved from
\url{http://dx.doi.org/10.12688/f1000research.2-191.v2}

Cranston, K., Harmon, L.J., O'Leary, M.A. \& Lisle, C. (2014). Best
practices for data sharing in phylogenetic research. \emph{PLoS Curr}.
Retrieved from
\url{http://dx.doi.org/10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645}

Drew, B.T., Gazis, R., Cabezas, P., Swithers, K.S., Deng, J., Rodriguez,
R., Katz, L.A., Crandall, K.A., Hibbett, D.S. \& Soltis, D.E. (2013).
Lost branches on the tree of life. \emph{PLoS Biol}, \textbf{11},
e1001636. Retrieved from
\url{http://dx.doi.org/10.1371/journal.pbio.1001636}

Hartig, O. (2012). An introduction to sPARQL and queries over linked
data. \emph{Web engineering} pp. 506--507. Springer Science + Business
Media. Retrieved from
\url{http://dx.doi.org/10.1007/978-3-642-31753-8_56}

Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
mapping of morphological characters. \emph{Systematic Biology},
\textbf{52}, 131--158. Retrieved from
\url{http://dx.doi.org/10.1080/10635150390192780}

Lang, D.T. (2013). \emph{XML: Tools for parsing and generating xML
within r and s-plus.} Retrieved from
\url{http://CRAN.R-project.org/package=XML}

Maddison, D., Swofford, D. \& Maddison, W. (1997). NEXUS: An extensible
file format for systematic information. \emph{Syst. Biol.}, \textbf{46},
590--621. Retrieved from
\url{http://www.ncbi.nlm.nih.gov/pubmed/11975335}

Midford, P., Dececchi, T., Balhoff, J., Dahdul, W., Ibrahim, N., Lapp,
H., Lundberg, J., Mabee, P., Sereno, P., Westerfield, M., Vision, T. \&
Blackburn, D. (2013). The vertebrate taxonomy ontology: A framework for
reasoning across model organism and species phenotypes. \emph{J. Biomed.
Semantics}, \textbf{4}, 34. Retrieved from
\url{http://dx.doi.org/10.1186/2041-1480-4-34}

NESCENT R Hackathon Team. (2014). \emph{Phylobase: Base package for
phylogenetic structures and comparative data}. Retrieved from
\url{http://CRAN.R-project.org/package=phylobase}

O'Meara, B. (2014). CRAN task view: Phylogenetics, especially
comparative methods. Retrieved from
\url{http://cran.r-project.org/web/views/Phylogenetics.html}

Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: Analyses of
phylogenetics and evolution in R language. \emph{Bioinformatics},
\textbf{20}, 289--290.

Parr, C.S., Guralnick, R., Cellinese, N. \& Page, R.D.M. (2011).
Evolutionary informatics: unifying knowledge about the diversity of
life. \emph{Trends in ecology \& evolution}, \textbf{27}, 94--103.
Retrieved from \url{http://www.ncbi.nlm.nih.gov/pubmed/22154516}

Pennell, M.W., Eastman, J.M., Slater, G.J., Brown, J.W., Uyeda, J.C.,
Fitzjohn, R.G., Alfaro, M.E. \& Harmon, L.J. (2014). Geiger v2.0: An
expanded suite of methods for fitting macroevolutionary models to
phylogenetic trees. \emph{Bioinformatics}, \textbf{30}, 2216--2218.

Piel, W.H., Chan, L., Dominus, M.J., Ruan, J., Vos, R.A. \& Tannen, V.
(2009). TreeBASE v. 2: A database of phylogenetic knowledge. Retrieved
from \url{http://www.e-biosphere09.org}

Piel, W.H., Donoghue, M.J. \& Sanderson, M.J. (2002). TreeBASE: A
database of phylogenetic information. \emph{The interoperable `catalog
of life'} (eds J. Shimura, K.L. Wilson \& D. Gordon), pp. 41--47.
Research report. National Institute for Environmental Studies, Tsukuba,
Japan. Retrieved from
\url{http://donoghuelab.yale.edu/sites/default/files/124_piel_shimura02.pdf}

Prud'hommeaux, E. (2014). SPARQL query language for rDF. \emph{W3C}.
Retrieved from \url{http://www.w3.org/TR/rdf-sparql-query/}

R Core Team. (2014). \emph{R: A language and environment for statistical
computing}. R Foundation for Statistical Computing, Vienna, Austria.
Retrieved from \url{http://www.R-project.org/}

Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
M.C. (2010). Data archiving. \emph{Evolution}, \textbf{64}, 603--604.
Retrieved from \url{http://dx.doi.org/10.1111/j.1558-5646.2009.00940.x}

Revell, L.J. (2012). Phytools: An r package for phylogenetic comparative
biology (and other things). \emph{Methods in Ecology and Evolution},
\textbf{3}, 217--223.

Stodden, V. (2014). The scientific method in practice: Reproducibility
in the computational sciences. \emph{SSRN Journal}. Retrieved from
\url{http://dx.doi.org/10.2139/ssrn.1550193}

Stoltzfus, A., O'Meara, B., Whitacre, J., Mounce, R., Gillespie, E.L.,
Kumar, S., Rosauer, D.F. \& Vos, R.A. (2012). Sharing and re-use of
phylogenetic trees (and associated data) to facilitate synthesis.
\emph{BMC Research Notes}, \textbf{5}, 574. Retrieved from
\url{http://dx.doi.org/10.1186/1756-0500-5-574}

Tenopir, C., Allard, S., Douglass, K., Aydinoglu, A.U., Wu, L., Read,
E., Manoff, M. \& Frame, M. (2011). Data sharing by scientists:
Practices and perceptions (C. Neylon, Ed.). \emph{PLoS ONE}, \textbf{6},
e21101. Retrieved from
\url{http://dx.doi.org/10.1371/journal.pone.0021101}

Vos, R.A., Balhoff, J.P., Caravas, J.A., Holder, M.T., Lapp, H.,
Maddison, W.P., Midford, P.E., Priyam, A., Sukumaran, J., Xia, X. \&
Stoltzfus, A. (2012). NeXML: Rich, extensible, and verifiable
representation of comparative data and metadata. \emph{Systematic
Biology}, \textbf{61}, 675--689. Retrieved from
\url{http://dx.doi.org/10.1093/sysbio/sys025}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{RNeXML: a package for reading and writing richly annotated phylogenetic,
character, and trait data in R}
    \author[ucb]{Carl Boettiger\corref{c1}}
   \ead{cboettig(at)gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[ropensci]{Scott Chamberlain}

@@

      \address[ucb]{University of California, Berkeley, 130 Mulford Hall \#3114, Berkeley,
CA 94720-3114, USA}    
    \address[ropensci]{University of California, Berkeley, CA, USA}    
    \address[NBC]{Naturalis Biodiversity Center, Leiden, the Netherlands}    
    \address[dukeplus]{Center for Genomic and Computational Biology, Duke University, and
National Evolutionary Synthesis Center, Durham, NC, USA}

@@

The exchange problem for phylogenetic data is particularly acute in
light of the challenges in finding and sharing phylogenetic data without
the otherwise common loss of most data and metadata semantics (Stoltzfus
\emph{et al.} 2012; Drew \emph{et al.} 2013; Cranston \emph{et al.}
2014). For example, the still popular NEXUS file format (Maddison
\emph{et al.} 1997) cannot consistently represent horizontal gene
transfer or ambiguity in reading a character (such as a DNA sequence
base pair). This and other limitations have led to modifications of
NEXUS in different ways for different needs, with the unfortunate result
that NEXUS files generated by one program can be incompatible with
another (Vos \emph{et al.} 2012). Without a formal grammar, software
based on NEXUS files may also make inconsistent assumptions about
tokens, quoting, or element lengths. Vos \emph{et al.} (2012) estimates
that as many as 15\% of the NEXUS files in the CIPRES portal contain
unrecoverable but hard to diagnose errors.

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in Vos \emph{et al.} (2012). In brief,
NeXML was designed with the following important properties. First, NeXML
is defined by a precise grammar that can be programmatically
\textbf{validated}; i.e., it can be verified whether a file precisely
follows this grammar, and therefore whether it can be read (parsed)
without errors by software that uses the NeXML grammar (e.g.~RNeXML) is
predictable. Second, NeXML is \textbf{extensible}: a user can define
representations of new, previously unanticipated information (as we will
illustrate) without violating its defining grammar. Third and most
importantly, NeXML is rich in \textbf{computable semantics}: it is
designed for expressing metadata such that machines can understand their
meaning and make inferences from it. For example, OTUs in a tree or
character matrix for frog species can be linked to concepts in a
formally defined hierarchy of taxonomic concepts such as the Vertebrate
Taxonomy Ontology (Midford \emph{et al.} 2013), which enables a machine
to infer that a query for amphibia is to include the frog data in what
is returned. (For a more broader discussion of the value of such
capabilities for evolutionary and biodiversity science we refer the
reader to Parr \emph{et al.} (2011).)

@@

The \texttt{RNeXML} package is written entirely in R and available under
a Creative Commons Zero public domain waiver. The current development
version can be found on Github at
\href{}{https://github.com/ropensci/RNeXML}, and the stable version can
be installed from the CRAN repository. \texttt{RNeXML} is part of the
rOpenSci project. Users of \texttt{RNeXML} are encouraged to submit bug
reports or feature requests in the issues log on Github, or the
phylogenetics R users group list at \texttt{r-sig-phylo@r-project.org}
for help. Vignettes with more detailed examples of specific features of
RNeXML are distributed with the R package and serve as a supplement to
this manuscript. Each of the vignettes can be found at
\href{}{http://ropensci.github.io/RNeXML/}.

@@

\begin{verbatim}
         log snout-vent length reef-dwelling
taxon_1              4.7532824             1
taxon_10             3.1373971             0
taxon_2             -2.7624146             0
taxon_3              2.1049413             0
taxon_4             -4.9504770             0
taxon_5              1.2714718             1
taxon_6              6.2593966             1
taxon_7              0.9099634             1
taxon_8             -3.2777799             0
taxon_9              2.0959433             1
\end{verbatim}

@@

\subsection{\texorpdfstring{Creating and populating \texttt{nexml}
objects}{Creating and populating nexml objects}}\label{creating-and-populating-nexml-objects}

@@

As described in Vos \emph{et al.} (2012), to encode metadata annotations
NeXML uses the ``Resource Description Framework in Annotations'' (RDFa)
(Prud'hommeaux 2014). This standard provides for a strict
machine-readable format yet enables future backwards compatibility with
compliant NeXML parsers (and thus \texttt{RNeXML}), because the capacity
of a tool to \emph{parse} annotations is not predicated on
\emph{understanding} the meaning of annotations it has not seen before.

@@

Third, \texttt{RNeXML} integrates with the R package \texttt{taxize}
(Chamberlain \& Szöcs 2013) to mitigate one of the most common obstacles
to reuse of phylogenetic data, namely the misspellings and inconsistent
taxonomic naming with which OTU labels are often fraught. The
\texttt{taxize\_nexml()} method in \texttt{RNeXML} uses \texttt{taxize}
to match OTU labels against the NCBI database, and, where a unique match
is found, it annotates the respective OTU with the matching NCBI
identifier.

@@

Because NeXML expresses all metadata using the RDF standard, and stores
them compliant with RDFa, they can be extracted as an RDF graph,
queried, analyzed, and mashed up with other RDF data, local or on the
web, using a wealth of off-the-shelf tools for working with RDF (see
Prud'hommeaux (2014) or Hartig (2012)). Examples for these possibilities
are included in the \texttt{RNeXML} SPARQL vignette (a recursive acronym
for SPARQL Protocol and RDF Query Language, see
\url{http://www.w3.org/TR/rdf-sparql-query/}), and the package also
comes with a demonstration that can be run from R using the following
command: \texttt{demo("sparql",\ "RNeXML")}).

@@

Data archiving is increasingly required by scientific journals,
including in evolutionary biology, ecology, and biodiversity (e.g.
Rausher \emph{et al.} (2010)). The effort involved with preparing and
submitting properly annotated data to archives remains a notable barrier
to the broad adoption of data archiving and sharing as a normal part of
the scholarly publication workflow (Tenopir \emph{et al.} 2011; Stodden
2014). In particular, the majority of phylogenetic trees published in
the scholarly record are inaccessible or lost to the research community
(Drew \emph{et al.} 2013).

@@

\hypertarget{refs}{}
\hypertarget{ref-zenodo}{}
Boettiger, C., Vos, R., Chamberlain, S. \& Lapp, H. (2014). RNeXML
v2.0.0. Retrieved from \url{http://dx.doi.org/10.5281/zenodo.13131}

\hypertarget{ref-Bollbackux5f2006}{}
Bollback, J. (2006).\emph{BMC Bioinformatics}, \textbf{7}, 88. Retrieved
from \url{http://dx.doi.org/10.1186/1471-2105-7-88}

\hypertarget{ref-Chamberlainux5f2013}{}
Chamberlain, S.A. \& Szöcs, E. (2013). Taxize: Taxonomic search and
retrieval in r. \emph{F1000Research}. Retrieved from
\url{http://dx.doi.org/10.12688/f1000research.2-191.v2}

\hypertarget{ref-Cranstonux5f2014}{}
Cranston, K., Harmon, L.J., O'Leary, M.A. \& Lisle, C. (2014). Best
practices for data sharing in phylogenetic research. \emph{PLoS Curr}.
Retrieved from
\url{http://dx.doi.org/10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645}

\hypertarget{ref-Drewux5f2013}{}
Drew, B.T., Gazis, R., Cabezas, P., Swithers, K.S., Deng, J., Rodriguez,
R., Katz, L.A., Crandall, K.A., Hibbett, D.S. \& Soltis, D.E. (2013).
Lost branches on the tree of life. \emph{PLoS Biol}, \textbf{11},
e1001636. Retrieved from
\url{http://dx.doi.org/10.1371/journal.pbio.1001636}

\hypertarget{ref-Hartigux5f2012}{}
Hartig, O. (2012). An introduction to sPARQL and queries over linked
data. \emph{Web engineering} pp. 506--507. Springer Science + Business
Media. Retrieved from
\url{http://dx.doi.org/10.1007/978-3-642-31753-8_56}

\hypertarget{ref-Huelsenbeckux5f2003}{}
Huelsenbeck, J.P., Nielsen, R. \& Bollback, J.P. (2003). Stochastic
mapping of morphological characters. \emph{Systematic Biology},
\textbf{52}, 131--158. Retrieved from
\url{http://dx.doi.org/10.1080/10635150390192780}

\hypertarget{ref-XML}{}
Lang, D.T. (2013). \emph{XML: Tools for parsing and generating xML
within r and s-plus.} Retrieved from
\url{http://CRAN.R-project.org/package=XML}

\hypertarget{ref-Maddisonux5f1997}{}
Maddison, D., Swofford, D. \& Maddison, W. (1997). NEXUS: An extensible
file format for systematic information. \emph{Syst. Biol.}, \textbf{46},
590--621. Retrieved from
\url{http://www.ncbi.nlm.nih.gov/pubmed/11975335}

\hypertarget{ref-Midford2013}{}
Midford, P., Dececchi, T., Balhoff, J., Dahdul, W., Ibrahim, N., Lapp,
H., Lundberg, J., Mabee, P., Sereno, P., Westerfield, M., Vision, T. \&
Blackburn, D. (2013). The vertebrate taxonomy ontology: A framework for
reasoning across model organism and species phenotypes. \emph{J. Biomed.
Semantics}, \textbf{4}, 34. Retrieved from
\url{http://dx.doi.org/10.1186/2041-1480-4-34}

\hypertarget{ref-phylobase}{}
NESCENT R Hackathon Team. (2014). \emph{Phylobase: Base package for
phylogenetic structures and comparative data}. Retrieved from
\url{http://CRAN.R-project.org/package=phylobase}

\hypertarget{ref-taskview}{}
O'Meara, B. (2014). CRAN task view: Phylogenetics, especially
comparative methods. Retrieved from
\url{http://cran.r-project.org/web/views/Phylogenetics.html}

\hypertarget{ref-Paradisux5f2004}{}
Paradis, E., Claude, J. \& Strimmer, K. (2004). APE: Analyses of
phylogenetics and evolution in R language. \emph{Bioinformatics},
\textbf{20}, 289--290.

\hypertarget{ref-Parr2011}{}
Parr, C.S., Guralnick, R., Cellinese, N. \& Page, R.D.M. (2011).
Evolutionary informatics: unifying knowledge about the diversity of
life. \emph{Trends in ecology \& evolution}, \textbf{27}, 94--103.
Retrieved from \url{http://www.ncbi.nlm.nih.gov/pubmed/22154516}

\hypertarget{ref-Pennell2014}{}
Pennell, M.W., Eastman, J.M., Slater, G.J., Brown, J.W., Uyeda, J.C.,
Fitzjohn, R.G., Alfaro, M.E. \& Harmon, L.J. (2014). Geiger v2.0: An
expanded suite of methods for fitting macroevolutionary models to
phylogenetic trees. \emph{Bioinformatics}, \textbf{30}, 2216--2218.

\hypertarget{ref-Pielux5f2009}{}
Piel, W.H., Chan, L., Dominus, M.J., Ruan, J., Vos, R.A. \& Tannen, V.
(2009). TreeBASE v. 2: A database of phylogenetic knowledge. Retrieved
from \url{http://www.e-biosphere09.org}

\hypertarget{ref-Pielux5f2002}{}
Piel, W.H., Donoghue, M.J. \& Sanderson, M.J. (2002). TreeBASE: A
database of phylogenetic information. \emph{The interoperable `catalog
of life'} (eds J. Shimura, K.L. Wilson \& D. Gordon), pp. 41--47.
Research report. National Institute for Environmental Studies, Tsukuba,
Japan. Retrieved from
\url{http://donoghuelab.yale.edu/sites/default/files/124_piel_shimura02.pdf}

\hypertarget{ref-W3Cux5f2014}{}
Prud'hommeaux, E. (2014). SPARQL query language for rDF. \emph{W3C}.
Retrieved from \url{http://www.w3.org/TR/rdf-sparql-query/}

\hypertarget{ref-R}{}
R Core Team. (2014). \emph{R: A language and environment for statistical
computing}. R Foundation for Statistical Computing, Vienna, Austria.
Retrieved from \url{http://www.R-project.org/}

\hypertarget{ref-Rausherux5f2010}{}
Rausher, M.D., McPeek, M.A., Moore, A.J., Rieseberg, L. \& Whitlock,
M.C. (2010). Data archiving. \emph{Evolution}, \textbf{64}, 603--604.
Retrieved from \url{http://dx.doi.org/10.1111/j.1558-5646.2009.00940.x}

\hypertarget{ref-Revellux5f2012}{}
Revell, L.J. (2012). Phytools: An r package for phylogenetic comparative
biology (and other things). \emph{Methods in Ecology and Evolution},
\textbf{3}, 217--223.

\hypertarget{ref-Stoddenux5f2014}{}
Stodden, V. (2014). The scientific method in practice: Reproducibility
in the computational sciences. \emph{SSRN Journal}. Retrieved from
\url{http://dx.doi.org/10.2139/ssrn.1550193}

\hypertarget{ref-Stoltzfusux5f2012}{}
Stoltzfus, A., O'Meara, B., Whitacre, J., Mounce, R., Gillespie, E.L.,
Kumar, S., Rosauer, D.F. \& Vos, R.A. (2012). Sharing and re-use of
phylogenetic trees (and associated data) to facilitate synthesis.
\emph{BMC Research Notes}, \textbf{5}, 574. Retrieved from
\url{http://dx.doi.org/10.1186/1756-0500-5-574}
